A Convolutional Neural Network for Outsole Recognition\tnoteref{t1}

### 2019-06-25 paper/paper.Rnw
All captions except heatmaps

if (grepl("!School", model_path)) {
  newest_model <- get_newest(dir = model_path, pattern = "weights.h5")
} else {
  newest_model <- list(path = file.path(model_path, "20190514-155453"),
      base_file = "2019-05-14_19:45:22_vgg16_onehotaug_9class_256-weights.h5",
      prefix = "vgg16_onehotaug_9class_256", start_date = "2019-05-14_19:45:22",
      process_dir = "20190514-155453")
}

@@

\section{Materials and Methods}
%\fix{We may want to mention some of the methods in the introduction for continuity sake...}
%\svp{I was talking about the methods above (e.g. fractal decomp, scale invariant feature recognition, etc., not methods we used)}
Ultimately, most pattern evidence is stored in photographic form; as a result, we must use image analysis methods to create numerical features from the original visual record. There are a number of methods that may be employed to identify shapes and features in an image, such as Fourier\mt{-Mellin} transforms, \mt{fractal decomposition, scale-invariant feature recognition}, Hough transformations \citep{ballardGeneralizingHoughTransform1981} and other low-level feature extraction methods aimed at detecting specific shapes, such as edges, corners, blobs, or ridges \citep[Ch 15]{machineVision}. While these methods are useful in identifying these specific features at a low level, they can be computationally intensive and only identify features on a very small scale; as a result, they cannot reliably identify large geometric shapes like those that may be found in a high resolution outsole image.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{images/vgg16-shoe-nolabel}
\caption[VGG16 structure]{VGG16 consists of five convolutional blocks that make up the model base. Each convolutional block contains an increasing number of increasingly complex features. After the convolutional blocks, the fully connected layers of the model head are used to make global connections between separate features.}\label{fig:VGG16-structure}
\end{figure}

@@

<<vgg16layers, fig.width = 7, fig.height = 5.75, fig.cap = "A selection of filters from the convolutional layers of VGG16.">>=

@@

<<overall-roc, fig.width = 5, fig.height = 5, out.width = ".5\\textwidth", fig.cap = "ROC curve showing overall model performance.", fig.scap = "Overall model performance ROC curve.">>=
library(pROC)
pred_df <- as_tibble(preds) %>% gather(key = feature, value = value)
test_labs_df <- as_tibble(test_labs) %>% gather(key = feature, value = value)
whole_model_roc <- roc(test_labs_df$value, pred_df$value)

@@

<<class-roc, fig.width = 8, fig.height = 6, out.width = "\\textwidth", fig.cap = "Class-by-class ROC curves. AUC is area under the curve, a measure of overall model performance. Equal error rates are marked, indicating the position at which there is equal probability of a false positive or false negative error.", fig.scap = "Class-by-class ROC curves.">>=
aucs <- plot_onehot_roc(preds, test_labs, str_to_title(classes))
thresholds <- purrr::map_dbl(aucs$data$eer, ~.$thresholds)
aucs$data$thresholds <- thresholds

@@

The first rows of \autoref{fig:exemplars} are images which contain only one of the nine classes. Most of the predictions for these images are promising; there are some false-positive labels that indicate confusions between circle/text and polygon/quad, but all of the true shapes are corectly identified by the model above EER. The next image belongs to the ``other" category, as it does not contain any of the nine classes; the model correctly assigns near-zero probabilities to each class. The final rows are of images that belong to more than one class category. Again, predictions are largely accurate, with most of the prominent shapes correctly identified above EER. \fix{Expand on specific misclassifications more, after determining which examples to delete.}

\fix{DC moved to heatmap, green block text could still be removed}
\fix{Write/check figure captions}

@@

<<exemplars, fig.cap = "Examples containing a single shape", fig.pos = "p!", out.width = "100%", cache = F, fig.width = 8, fig.height = 12>>=
single_exemplars <- list.files(file.path(wd, "images", "consistency",
                                         "single-exemplar"), full.names = T)
multi_exemplars <- list.files(file.path(wd, "images", "consistency",
                                        "multi-exemplar"), full.names = T)
pred_prob_plot(rev(c(single_exemplars, multi_exemplars)),
               loaded_model, sort = F, eer = eers)
@

While figure \ref{fig:exemplars} shows representative examples of classifications across multiple shape categories, it is also useful to examine many cases within a given shape to get a sense of model consistency. \autoref{fig:circle-text-consistency} shows a number of images containing circles and/or text. \fix{The top four rows show text contained within a separate, distinct circle shape, and the next five rows show circles represented as the letter ``o" in text. The remaining images contain only one of either text or circles, as well as other possible shape categories.} In particular, the first two of these images contain text that use the capital letter ``G", the next two images contain no circular shapes, and the final \fix{N} images contain pure circles that do not appear to belong to text. The model is able to identify the circle in all cases where a circle is truly present, regardless of whether the circle is part of the text or it is distinct; however, the circle prediction is also strong in three of the four images that do not contain a circle, which indicates that the model is having dificulty distinguishing the two categories. While the circle predictions are above the EER for letters such as C and S, these predictions are generally much lower than those where distinct circles or Os are seen, indicating that there are some indications that features are not as strong in the quasi-circle shapes, but that the model is not able to completely separate Os from C, S, and G shapes which are also very circular. This confusion is complicated by the frequency of images which contain both circles and text: text is present in \Sexpr{100 * round(sum(as.logical(grepl("circle", dfunion$name) & grepl("text", dfunion$name))) / sum(grepl("circle", dfunion$name)), 2)}\% of our labeled images containing circles, and circles are present in \Sexpr{100 * round(sum(as.logical(grepl("circle", dfunion$name) & grepl("text", dfunion$name))) / sum(grepl("text", dfunion$name)), 2)}\% of our images containing text. In addition, the distinction between a circle and an ``o" comes from the context around the shape, which is much more difficult for the model to parse than simply detecting a shape. More importantly, this behavior does not suggest poor model performance as much as a weakness of the classification scheme, which will be discussed in a later section.

% \fix{make sure text matches figure!}

<<circle-text-consistency, fig.cap = "Text with circles or circle-like shapes", fig.pos = "p!", out.width = "100%", cache = T, fig.width = 8, fig.height = 12>>=
circle_text <- list.files(file.path(wd, "images", "consistency", "circle-text"), full.names = T)
pred_prob_plot(rev(circle_text), loaded_model, sort = F, eer = eers)
@

@@

<<low-contrast-consistency, fig.cap = "Images with low contrast", fig.pos = "p!", out.width = "100%", cache = F>>=
lowcontrast <- list.files(file.path(wd, "images", "consistency", "low-contrast2"), full.names = T)
pred_prob_plot(lowcontrast, loaded_model, sort = F, eer = eers)
@

@@

knitr::include_graphics(heatmaps[grepl("Ugg-grey", heatmaps)])
knitr::include_graphics(heatmaps[grepl("text", heatmaps)])
knitr::include_graphics(heatmaps[grepl("polygon_recolor", heatmaps)])
knitr::include_graphics(heatmaps[grepl("corks", heatmaps)])

@@

% Discussion of weaknesses
Although model predictions are generally accurate, there are a number of systematic misclassifications. The model has substantial difficulty predicting images with low contrast because image features are obscured. Exploring methods of color correction, such as histogram equalization, may prove useful for eliminating the effect of contrast on predictions. \svp{Initial attempts at implementing histogram equalization suggest that the overall accuracy improvement is modest; the model fit to images which have been equalized had approximately 92.5\% validation accuracy, an improvement of 1.5\% over the default model described in this paper.}

Another source of systematic misclassifications stems from the inherent overlap between features of common geometric shapes\svp{. }CNNs make predictions via bottom-up feature integration; \svp{as a result,} it is more difficult for the model to distinguish between geometrically similar shapes, such as chevrons and triangles. This issue is exacerbated by high variability of some shapes \svp{commonly found in outsole patterns}, which requires the model to form a flexible understanding of each shape category \svp{and makes} distinction between similar shapes even more difficult. These issues of overlap and variability are at least in part due to our decision to use the features used by practitioners directly; some of this may also result from our operational definitions of various shapes, which \svp{were created before we had a complete understanding of the practical limitations of CNNs.\sout{did not take into account the model's capabilities and initial training.}} While a \fix{full} remedy may not be possible, there may be some modifications to operational class definitions that we can make to optimize labels for model prediction. \svp{In addition, we will continue to improve the labeling of shapes in our database with the goal of expanding the breadth and quantity of labeled images available for model training.}

\svp{Although there are many areas for potential improvement in future iterations of this model, we have demonstrated the effectiveness of transfer learning to provide an initial label for shoe outsole patterns. As} most of the misclassifications are systematic, it is clear that the model is able to learn features and apply rules consistently. \svp{We expect that future modifications to the labeling criteria and data processing pipeline will improve the model's accuracy, but the 91\% validation accuracy} \fix{really should be looking at test set accuracy, but I don't know that number off the top of my head...} \svp{is promising considering that we do not have estimates of the consistency of examiners' estimates on ambiguous features.}  \svp{Taken together, these factors } show that \svp{Convolutional Neural Networks} are a useful tool for classifying a large number of outsoles within a common classification scheme and can ultimately help streamline the current process of sorting and comparing shoes by class characteristics.

% \fix{Although the model is not perfect in its current form, as a first pass tool it is performing well and improvements seem promising. That most of the misclassifications are systematic shows that the model is able to learn features and apply rules consistently. There are possible modifications to the labeling scheme that may make model predictions align more closely to human-assigned labels, but model consistency is a much higher good than ``accuracy" (in the sense of agreeing with human-assigned labels). Examiners themselves may vary in how to label ambiguous features (e.g., the half-rounded quadrilaterals in \fix{the dc heatmap}), so a consistent classification method provides a more unified definition of certain geometric shapes and facilitates comparison of outsoles with similar geometric features. All in all, this model shows that CNNS are a useful tool for classifying a large number of outsoles within a common classification scheme and can ultimately help streamline the current process of sorting and comparing shoes by class characteristics.}
%
% \mt{I think the conclusion currently says everything I want it to (cutting off a conv block doesn't seem to help make the point). Definitely can use some language edits.}

>>>>

if (grepl("!School", model_path)) {
  newest_model <- get_newest(dir = model_path, pattern = "weights.h5")
  newest_model$prefix = "vgg16_onehotaug_9class_256"
  newest_model$start_date = "2019-05-14_19_45_22"
} else {
  newest_model <- list(path = file.path(model_path, "20190514-155453"),
      base_file = "2019-05-14_19:45:22_vgg16_onehotaug_9class_256-weights.h5",
      prefix = "vgg16_onehotaug_9class_256", start_date = "2019-05-14_19:45:22",
      process_dir = "20190514-155453")
}

@@

\section{Materials and Methods}
%\fix{We may want to mention some of the methods in the introduction for continuity sake...
Ultimately, most pattern evidence is stored in photographic form; as a result, we must use image analysis methods to create numerical features from the original visual record. There are a number of methods that may be employed to identify shapes and features in an image, such as Fourier\mt{-Mellin} transforms, \mt{fractal decomposition, scale-invariant feature recognition}, Hough transformations \citep{ballardGeneralizingHoughTransform1981} and other low-level feature extraction methods aimed at detecting specific shapes, such as edges, corners, blobs, or ridges \citep[Ch 15]{machineVision}. While these methods are useful in identifying these specific features at a low level, they can be computationally intensive and only identify features on a very small scale; as a result, they cannot reliably identify large geometric shapes like those that may be found in a high resolution outsole image.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{images/vgg16-shoe-nolabel}
\caption[VGG16 structure]{VGG16 consists of five convolutional blocks that make up the model base. Each convolutional block contains an increasing number of increasingly complex \mt{feature-detecting filters}. After the convolutional blocks, the fully connected layers of the model head are used to make global connections between separate features.}\label{fig:VGG16-structure}
\end{figure}

@@

<<vgg16layers, fig.width = 7, fig.height = 5.75, fig.cap = "{Visualizing convolutional filters requires generating images which maximally activate specific filters. This figure contains images that maximally activate a selection of filters from the convolutional blocks of VGG16. Filters from later blocks detect more complex features than those in earlier blocks. Much like human vision, a hierarchical structure of feature detection with increasing complexity allows CNN ``vision\" trained on a general task to be applied to new domains.}">>=

@@

<<overall-roc, fig.width = 5, fig.height = 5, out.width = ".5\\textwidth", fig.cap = "ROC curve showing overall model performance. \\mt{The curve indicates that model accuracy is generally high.}", fig.scap = "Overall model performance ROC curve.">>=
library(pROC)
pred_df <- as_tibble(preds) %>% gather(key = feature, value = value)
test_labs_df <- as_tibble(test_labs) %>% gather(key = feature, value = value)
whole_model_roc <- roc(test_labs_df$value, pred_df$value)

@@

<<class-roc, fig.width = 8, fig.height = 6, out.width = "\\textwidth", fig.cap = "Class-by-class ROC curves. AUC is area under the curve, a measure of overall model performance. Equal error rates (EER) are marked, indicating the position at which there is equal probability of a false positive or false negative error.", fig.scap = "Class-by-class ROC curves.">>=
aucs <- plot_onehot_roc(preds, test_labs, str_to_title(classes))
thresholds <- purrr::map_dbl(aucs$data$eer, ~.$thresholds)
aucs$data$thresholds <- thresholds

@@

The first rows of \autoref{fig:exemplars} are images which contain only one of the nine classes. Most of the predictions for these images are promising; there are some false-positive labels that indicate confusions between circle/text and polygon/quad, but all of the true shapes are corectly identified by the model above their respective EER. The next image belongs to the ``other" category, as it does not contain any of the nine classes; the model correctly assigns near-zero probabilities to each class. The final rows are of images that belong to more than one class category. Again, predictions are largely accurate, with most of the prominent shapes correctly identified above EER. \mt{The most notable misclassifications in these images are the false positive predictions of triangle in the image with green text and lines in the Adidas logo. Although the Adidas logo is technically made up of adjacent quadrilaterals, the prediction of a triangle is reasonable when considering the proximity of those quadrilaterals and the shape's features as a whole, and the relatively low predicted probability suggests that the model recognizes that this shape is ambiguous. All other predictions of these images are reasonably correct given our operational definition of class categories.}

@@

<<exemplars, fig.cap = "\\mt{Model predictions of images containing a single shape, no shape, or multiple shapes across all categories. Although there are some confusions (such as between circle and text in the single class images, and the false positive prediction of triangle in the image with green text), predictions are generally reasonable and consistent with the human-assigned labels of these images.}", fig.pos = "p!", out.width = "100%", cache = T, fig.width = 8, fig.height = 12>>=
single_exemplars <- list.files(file.path(wd, "images", "consistency",
                                         "single-exemplar"), full.names = T)
multi_exemplars <- list.files(file.path(wd, "images", "consistency",
                                        "multi-exemplar"), full.names = T)
pred_prob_plot(rev(c(single_exemplars, multi_exemplars)),
               loaded_model, sort = F, eer = eers)
@

While figure \ref{fig:exemplars} shows representative examples of classifications across multiple shape categories, it is also useful to examine many cases within a given shape to get a sense of model consistency. \autoref{fig:circle-text-consistency} shows a number of images containing circles and/or text. The top four rows show text contained within a separate, distinct circle shape, and the next five rows show circles represented as the letter ``o" in text. The remaining images contain only one of either text or circles, as well as other possible shape categories. In particular, the first two of these images contain text that use the capital letter ``G", the next two images contain no circular shapes, and the final 3 images contain pure circles that do not appear to belong to text. The model is able to identify the circle in all cases where a circle is truly present, regardless of whether the circle is part of the text or it is distinct; however, the circle prediction is also strong in three of the four images that do not contain a circle, which indicates that the model is having dificulty distinguishing the two categories. While the circle predictions are \mt{often} above the EER for letters such as G and S, these predictions are generally much lower than those where distinct circles or Os are seen, indicating that there are some indications that features are not as strong in the quasi-circle shapes, but that the model is not able to completely separate Os from C, S, and G shapes which are also very circular. This confusion is complicated by the frequency of images which contain both circles and text: text is present in \Sexpr{100 * round(sum(as.logical(grepl("circle", dfunion$name) & grepl("text", dfunion$name))) / sum(grepl("circle", dfunion$name)), 2)}\% of our labeled images containing circles, and circles are present in \Sexpr{100 * round(sum(as.logical(grepl("circle", dfunion$name) & grepl("text", dfunion$name))) / sum(grepl("text", dfunion$name)), 2)}\% of our images containing text. In addition, the distinction between a circle and an ``o" comes from the context around the shape, which is much more difficult for the model to parse than simply detecting a shape. More importantly, this behavior does not suggest poor model performance as much as a weakness of the classification scheme, which will be discussed in a later section.

<<circle-text-consistency, fig.cap = "\\mt{Model predictions of images containing circles and/or text. The model correctly identifies circles in all images where circles are present, but also predicts moderate probabilities of circles in many images where text is present with curved letters, such as ``G\". Text is correctly identified in most images, with the exception of the false negative prediction of the low-contrast Teva image and the false positive prediction of concentric circles. In general, model predictions are reasonable but suggest a few sources of systematic misclassification, which are discussed in a later section.}", fig.pos = "p!", out.width = "100%", cache = T, fig.width = 8, fig.height = 12>>=
circle_text <- list.files(file.path(wd, "images", "consistency", "circle-text"), full.names = T)
pred_prob_plot(rev(circle_text), loaded_model, sort = F, eer = eers)
@

@@

<<low-contrast-consistency, fig.cap = "\\mt{Model predictions for images with low contrast. While these images do truly contain a variety of shapes, the diffuse probabilities assigned by the model across all categories indicate that features are obscured by low contrast, which propogates through model training and prediction.}", fig.pos = "p!", out.width = "100%", cache = T>>=
lowcontrast <- list.files(file.path(wd, "images", "consistency", "low-contrast2"), full.names = T)
pred_prob_plot(lowcontrast, loaded_model, sort = F, eer = eers)
@

@@

knitr::include_graphics(heatmaps[grepl("Ugg-grey", heatmaps)])
knitr::include_graphics(heatmaps[grepl("text-seychelles", heatmaps)])
knitr::include_graphics(heatmaps[grepl("polygon_recolor", heatmaps)])
knitr::include_graphics(heatmaps[grepl("corks", heatmaps)])

@@

% Discussion of weaknesses
Although model predictions are generally accurate, there are a number of systematic misclassifications. The model has substantial difficulty predicting images with low contrast because image features are obscured. Exploring methods of color correction, such as histogram equalization, may prove useful for eliminating the effect of contrast on predictions. Initial attempts at implementing histogram equalization suggest that the overall accuracy improvement is modest; the model fit to images which have been equalized had approximately 92.5\% validation accuracy, an improvement of 1.5\% over the default model described in this paper.

Another source of systematic misclassifications stems from the inherent overlap between features of common geometric shapes. CNNs make predictions via bottom-up feature integration; as a result, it is more difficult for the model to distinguish between geometrically similar shapes, such as chevrons and triangles. This issue is exacerbated by high variability of some shapes commonly found in outsole patterns, which requires the model to form a flexible understanding of each shape category and makes distinction between similar shapes even more difficult. These issues of overlap and variability are at least in part due to our decision to use the features used by practitioners directly; some of this may also result from our operational definitions of various shapes, which were created before we had a complete understanding of the practical limitations of CNNs. While a complete remedy may not be possible, there may be some modifications to operational class definitions that we can make to optimize labels for model prediction. In addition, we will continue to improve the labeling of shapes in our database with the goal of expanding the breadth and quantity of labeled images available for model training.

Although there are many areas for potential improvement in future iterations of this model, we have demonstrated the effectiveness of transfer learning to provide an initial label for shoe outsole patterns. As most of the misclassifications are systematic, it is clear that the model is able to learn features and apply rules consistently. We expect that future modifications to the labeling criteria and data processing pipeline will improve the model's accuracy, but the 91\% validation accuracy \fix{really should be looking at test set accuracy, but I don't know that number off the top of my head...}\mt{See chunk below} is promising considering that we do not have estimates of the consistency of examiners' estimates on ambiguous features. Taken together, these factors show that Convolutional Neural Networks are a useful tool for classifying a large number of outsoles within a common classification scheme and can ultimately help streamline the current process of sorting and comparing shoes by class characteristics.

<<test-accuracy, echo = F>>=
# This is my stab at calculating test accuracy.
# .786 doesn't seem too promising :(

disc_preds <- sweep(preds, 2, eers) > 0 #predicted values minus EER, column wise
test_accuracy <- sum(disc_preds == test_labs) / (length(test_labs))
@
